Black-hole ringdown as a probe of higher-curvature gravity theories

### 2022-03-30 papers/paper_alt_theor_bounds/paper_alt_theor_bounds.tex
hs: still editing edgb section

\subsection{scalar-Gauss-Bonnet gravity}
%
This theory is given by the following action,
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \mathscr{G}
    \right],
\end{equation}
%
where $R$ is the Ricci scalar associated with the metric $g_{\alpha\beta}$ and
%$\varphi$ is a dynamical scalar field which couples to the Gauss-Bonnet
invariant $\mathscr{G}$,
%
\begin{equation} \label{eq:def_gb}
    \mathscr{G} =
    R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
    - 4 R^{\mu\nu}R_{\mu\nu}
    + R^2,
\end{equation}
%
with strength set by the dimensionful coupling constant $\ell_{\rm GB}$ with dimensions
of length. Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH solutions.
%
In the first class, the first derivative of the coupling
function $f'(\varphi) = \dd f  / \dd \varphi$ is always nonzero and BHs
are known to always support (secondary) scalar hair.
%
Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
$f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, one can show that theory admits the same stationary,
asymptotically flat BH solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the quadratic $f \propto \varphi^2$~\cite{Silva:2017uqg}
and Gaussian $f \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} couplings.
%
Here we will consider only the dilatonic theory.

@@

Let us contrast our results with those of Ref.~\cite{Lyu:2022gdr} which placed
the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the inspiral portion of the neutron star-black hole binaries GW200105 and GW200115~\cite{LIGOScientific:2021qlt} or
of Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, which placed the bound $\ell_{\rm GB} \leqslant 1.7$~km, by considering
a selection of black hole binaries from the GWTC-1 and GWTC-1 catalogs~\cite{}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and take into consideration only the leading-order
scalar field contribution arising from the dilatonic coupling.}

@@

@@

>>>>

\subsection{scalar-Gauss-Bonnet gravity}
%
This theory is given by the following action,
%
\begin{equation} \label{eq:action_sgb}
    S_{\rm GB} = \frac{1}{16 \pi}
    \int \dV
    \left[
    R - \tfrac{1}{2}(\nabla \varphi)^2
    + \tfrac{1}{4} \ell^{2}_{\rm GB} f(\varphi) \mathscr{G}
    \right],
\end{equation}
%
where $R$ is the Ricci scalar associated with the metric $g_{\alpha\beta}$ and
$\varphi$ is a dynamical scalar field which couples to the Gauss-Bonnet
invariant $\mathscr{G}$,
%
\begin{equation} \label{eq:def_gb}
    \mathscr{G} =
    R^{\mu\nu\rho\sigma}R_{\mu\nu\rho\sigma}
    - 4 R^{\mu\nu}R_{\mu\nu}
    + R^2,
\end{equation}
%
with strength set by the dimensionful coupling constant $\ell_{\rm GB}$ with dimensions
of length. Different subclasses of this theory are determined by the function $f(\varphi)$
and they can be classified into two classes based on the properties of their BH solutions.
%
In the first class, the first derivative of the coupling
function $f'(\varphi) = \dd f  / \dd \varphi$ is always nonzero and BHs
are known to always support (secondary) scalar hair.
%
Examples include the shift-symmetric $f \propto \varphi$ and dilatonic
$f \propto \exp(\varphi)$ couplings.
%
In the second class, $f'(\varphi) = 0$ can vanish for some constant $\varphi_0$.
%
In this case, one can show that theory admits the same stationary,
asymptotically flat BH solution of GR and scalarized BHs~\cite{Doneva:2017bvd,Silva:2017uqg,Dima:2020yac,Herdeiro:2020wei,Berti:2020kgk}.
%
Examples include the quadratic $f \propto \varphi^2$~\cite{Silva:2017uqg}
and Gaussian $f \propto \exp(-\varphi^2)$~\cite{Doneva:2017bvd} couplings.
%
Here we will consider only the dilatonic theory.

@@

@@

Let us contrast our results with those of Ref.~\cite{Lyu:2022gdr} which placed
the bound
%
$\ell_{\rm GB} \leqslant 1.33$~km
%
using the \emph{inspiral} portion of the neutron star-black hole binaries GW200105 and GW200115~\cite{LIGOScientific:2021qlt} or
of Refs.~\cite{Nair:2019iur,Perkins:2021mhb}, which placed the bound $\ell_{\rm GB} \leqslant 1.7$~km, by considering
a selection of black hole binaries from the GWTC-1 and GWTC-1 catalogs~\cite{}\footnote{These bounds
are, strictly speaking, valid only when the scalar field $\varphi$ is small, i.e., $\varphi \ll 1$, and
take into consideration only the leading-order scalar field contribution arising from the dilatonic coupling.}
%
In this cases

Hence, here we have an example of a theory in which the inspiral portion of the signal
is more constraining then the ringdown portion of the signal.
%
As we will next, the situation is the opposite for dCS gravity.

@@

\hs{TODO: say that we could try to expand the waveform model to include modification to
the inspiral part too; maybe through FTI~\cite{Mehta:2022pcn}?}

\hs{TODO: add that we could try to parametrize the final black hole mass and spin estimates formulas we
use at the moment~\cite{Hofmann:2016yih}. Maybe comment that a possible first attempt could follow~\cite{Buonanno:2007sv},
as done in~\cite{Jai-akson:2017ldo,Li:2020wse}. Say that ultimately we will need to informed by NR in beyond-GR.}

### 2022-03-30 papers/paper_alt_theor_bounds/sec_pe.tex
hs: still editing edgb section

\subsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using the Bayes
theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{p(\bm{\lambda} \vert \mathcal{H}) \, \mathcal{L}(d \vert \bm{\lambda},\mathcal{H})}{E(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$p(\bm{\lambda} \vert \mathcal{H})$ the prior,
$\mathcal{L}(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$E(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is a union of the GR waveform
model parameters~$\bm{\theta}$ (c.f.  Sec.~\ref{sec:review_pSEOB}), and
$\ell$,  the only non-GR parameter in this problem which, we recall, set
the characteristic length-scale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}. Hence,
%
\begin{equation}
\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
\label{eq:def_params}
\end{equation}

@@

Assuming stationary Gaussian noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln \mathcal{L}(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}\,,
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters (discussed further in the subsequent section), we stochastically
sample over the parameter space using a Markov-Chain Monte Carlo algorithm as implemented in
\texttt{LALInferenceMCMC}~\cite{Rover:2006ni,vanderSluys:2008qx},
a package part of the \texttt{LALInference} software suite~\cite{Veitch:2014wba,lalsuite}.
%
We subsequently marginalize over the remaining parameters to obtain the
posterior probability distribution function~(PDF) on $\ell$,  $P_j(\ell \vert d_j,\mathcal{H})$, our main parameter of interest.
%
% An example of this is shown in Fig. XX.

>>>>

\subsection{Bayesian formalism}
If we assume that a GW signal observed in detector data $d$ is accurately
described by our waveform model \pSEOB{}, we can infer the parameters of the
model, $\bm{\lambda}$, given the hypothesis $\mathcal{H}$, using Bayes' theorem,
%
\begin{equation}
P(\bm{\lambda} \vert d, \mathcal{H}) =
\frac{p(\bm{\lambda} \vert \mathcal{H}) \, \mathcal{L}(d \vert \bm{\lambda},\mathcal{H})}{E(d \vert \mathcal{H})}\,,
\label{eq:bayes}
\end{equation}
%
where $P(\bm{\lambda} \vert d, \mathcal{H})$ is the posterior probability distribution,
$p(\bm{\lambda} \vert \mathcal{H})$ the prior,
$\mathcal{L}(d \vert \bm{\lambda},\mathcal{H})$ the likelihood, and
$E(d \vert \mathcal{H})$ the evidence.
%
The set of parameters, $\bm{\lambda}$ is a union of the GR waveform
model parameters~$\bm{\theta}$ (c.f.  Sec.~\ref{sec:review_pSEOB}), and
$\ell$,  the only non-GR parameter in this problem which, we recall, set
the characteristic length-scale in which deviations from
GR become relevant in each of the theories described in
Sec.~\ref{sec:review_theories}. Hence,
%
\begin{equation}
\bm{\lambda} = \{\ell\} \cup \{\bm{\theta}\}.
\label{eq:def_params}
\end{equation}

@@

Assuming stationary Gaussian noise, we can write the (log) likelihood function as,
%
\begin{equation}
\ln \mathcal{L}(d \vert \bm{\lambda},\mathcal{H}) \propto
- \tfrac{1}{2}
\langle
d - h(\bm{\lambda}) \vert d - h(\bm{\lambda})
\rangle\,,
\end{equation}
%
with noise-weighted inner product $\langle \cdot | \cdot \rangle$ defined as,
%
\begin{equation}
\langle A | B \rangle =
\int_{f_{\rm low}}^{f_{\rm high}} \, \dd f \,
\frac{\tilde{A}^{\ast}(f) \tilde{B}(f) + \tilde{A}(f) \tilde{B}^{\ast}(f)}{S_{n}(f)}\,,
\end{equation}
%
where $\tilde{A}(f)$ is the Fourier transform of $A(t)$, the asterisk denotes
complex conjugation and $S_{n}(f)$ is the power spectrum density of the
detector.
%
Assuming a specific prior distribution for our parameters (discussed further in the subsequent section), we stochastically
sample over the parameter space using a Markov-Chain Monte Carlo algorithm as implemented in
\texttt{LALInferenceMCMC}~\cite{Rover:2006ni,vanderSluys:2008qx},
a package part of the \texttt{LALInference} software suite~\cite{Veitch:2014wba,lalsuite}.
%
We subsequently marginalize over the remaining parameters to obtain the
posterior probability distribution function~(PDF) on $\ell$,  i.e., $P_j(\ell \vert d_j,\mathcal{H})$,
our main parameter of interest.
